Hilbert space methods to approximate Gaussian processes using Stan

### 2019-03-10 Paper/manuscript/manuscript_05.tex
improve Section 3

Implementation and performance of the HSGP model depends on the number of basis functions and boundary settings, which are, at the same time, related  to each other. The time of computation in the approximation mainly depends on the number of basis functions. Furthermore, the appropriate values for the number of basis functions and boundary depend on the non-linear functional effects of the function characterized by the lengthscale of the covariance function.

In this section we analyze the effects of the number of basis functions and boundary on the approximation accuracy. We present recommendations how they should be chosen and diagnostics to check the accuracy of the approximation after the inference.
We illustrate graphically how the number of basis functions, boundary and lengthscale relate to each other and affect the accuracy of the approximation.
% This graph basically provides the recommended values for these factors in order to get a close approximation. This graph also allows us to diagnose an actual fit by checking whether the value of the estimated lengthscale does not keep to those that the Figure brings as suitable ones.

We compare regular GP and HSGP fitted for different dataset with different "wigglyness".
%The comparison is made through comparing their estimated lengthscales.

@@

The approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, e.g. in a 1-$D$ input space $x\in \Omega=[-L,L]\subset \mathbb{R}$:
%
\begin{equation}\label{diffcov}
k(\tau) = \sum_{j}^{\infty} S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}

\noindent where $L$ is the boundary, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are those in equation (\ref{bf&lambda}) for the unidimensional case and those in equation (\ref{bf&lambda_multi}) for the multidimensional case.

The approximation to the covariance function becomes exact in the limit when the number of basis functions approach infinity, $j=1,\cdots,\infty$. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions be less than a threshold $\epsilon$,
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau)d(\tau) - \int \sum_{j}^m S_{\theta}(\sqrt{\lambda_j}) \phi_j(\tau) \phi_j(0) d(\tau) < \epsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy eq. (\ref{diff_covs}) depends on the non-linear effects of the function to be learned, that is, on its lengthscale $\ell$. The approximation also depends on boundary $L$ (equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})) which will affect the performance specially near the boundaries. The boundary will also have an effect on the number of basis functions needed in the approximation. The boundary $L$ is set as an extension of the desired predicting input domain $\Psi$. Considering a zero-mean input domain $\Psi=[-S,S] \in \mathbb{R}$, the boundary is defined as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

\noindent where $S$ represents the half-range of the input space, and $c$ ($c \ge 1$) is the proportional extension factor. From now on we are going to refer to this proportional factor $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by the half-range $S$ of the input space.

For a naive illustration of the effects of these two factors, number of basis functions $m$ and boundary $L$, on the performance of the approximation, a set of noisy observations are drawn from a GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$, and input vales within the zero-mean input domain with half-range $S=1$. HSGP models with different number of basis functions and boundaries are fitted to this data. The lengthscale and marginal variance parameters are considered fixed to the true values of the data-generative GP model. Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of the number of basis functions and the boundary, respectively, on the approximation of the covariance function. On the one hand, for a fixed $L$, let's suppose it is set to a right choice, Figure \ref{fig1_Post_J} shows the effect of changing the number of basis functions in approaching the posterior mean and the covariance function. It can be seen how the number of basis functions affects the non-linear effects of the posteriors. Fewer basis functions implies smoother covariance functions and consequently smoother posteriors. The more "wigglyness" of the function, the more basis functions are needed. On the other hand, and similarly, for a fixed number of basis functions $m$, let's suppose it is set to a right choice, Figure \ref{fig2_Post_L} shows the effect of changing the boundary $L$ in approaching the posterior mean and the covariance function. It can be seen how the box size mainly affects the approximation near the boundaries of the function and affects the covariances at long distances. \\

@@

In the previous Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} we focused on illustrating the individual effects of the number of basis functions and the boundary. Furthermore, for a clearer illustration of these effects, the lengthscale and marginal variance parameters in the HSGP model were fixed to their true values. Now we are going to focus on analyzing the interaction effects between these two factors, number of basis functions and the boundary factor, on the performance of the approximation. The lengthscale and marginal variance will not be fixed rather they will be estimated either in regular and HSGP models.  
%The right-side plots in Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} are computed using the lengthscale and magnitud values of the true GP prior. Now, we focus on the estimated lengthscale and magnitud parameters after fitting the model, either of the regular GP model fit and approximate GP model fit. 
Figure \ref{fig4_Post_&_Cov_part1} shows the posteriors means and the covariance functions obtained after fitting the data, for different number of basis functions and boundary factor. Figure \ref{fig5_MSE_vs_J} shows the root mean square error (RMSE) of the approximate GP model, computed against the regular GP model, as a function of the number of basis functions and the boundary factor. Figure \ref{fig6_lscale_vs_J} shows the lengthscale and marginal variance for the regular GP model and for the HSGP models as a function of the number of basis functions and boundary factor. Looking at the RMSEs in Figure \ref{fig5_MSE_vs_J}, it can be drawn that the optimal choice in terms of precision and computations would be 15 basis functions and a boundary factor between 1.5 and 2.5. The choice of 10 basis functions and a boundary factor of 1.5 could be an accurate enough choice. This same conclusion can be also be intuitively seen in the posterior mean and covariance function plots in Figure \ref{fig4_Post_&_Cov_part1}.

From Figure \ref{fig5_MSE_vs_J}, a behaviour in terms of performance of the HSGP model as a function of the number of basis functions and the boundary factor, can be deduced: \\
$\bullet$ as the boundary factor increases, more basis functions are needed,
\\
$\bullet$ as fewer basis functions are used, the boundary factor must decrease.

Similar conclusions can be stated from results of Figure \ref{fig6_lscale_vs_J}. This Figure shows the estimated lengthscale and marginal variance as function of the number of basis functions and boundary factor. In addition to the above conclusions, in Figure \ref{fig6_lscale_vs_J} it can be easily seen that exists a minimum value for the boundary factor under which a close approximation will never be achieved. \\

@@

Additionally, there exit a relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the function. Figure \ref{fig3_lscale_vs_J_vs_c_part1} collects how these three factors relate each other in relation to the performance of the approximation evaluated using eq. (\ref{diff_covs}). On the X-axis of this plot is placed the lengthscale of the process normalized by the half-range of the data $\ell/S$. The countour lines gather the boundary factor $c$. And the Y-axis represents the number of basis functions. Thus, this figure give us, for a certain GP function with lengthscale $\ell$ and given a boundary factor $c$, the minimum number of basis functions needed to achieve a close approximation in terms of satisfying eq. (\ref{diff_covs}). Alternatively, this figure could also be read as the minimum boundary factor $c$ that we should use given a number of basis functions, for certain GP function with certain lengthscale. And also this figure could also be read as the minimum functional lengthscales that can be closely approximated given a number of basis functions and a boundary factor. This figure basically collects the behaviour of the HSGP model as a function of the lengthscale, number of basis functions and boundary factor:

$\bullet$ As the lengthscale of the function increases, the boundary factor and the number of basis functions needed for the approximation decrease.

$\bullet$ There is a minimum boundary factor under which a close approximation never is going to be achieved (The contour lines have an end in function of the lengthscale). This statement match with one of the behaviour recognized in Figure \ref{fig6_lscale_vs_J} where were a minimum value for the boundary factor under which a close approximation will never be achieved.

$\bullet$ The lower the boundary factor, the fewer basis functions and the smaller lengthscales that are suitable to be fitted.

$\bullet$ This plot serves as a diagnosis tool in the sense that if a estimated lengthscale is lower than that minimum that the figure bring us, given certain number of basis functions and certain boundary factor, means that the number of basis functions should be increased or the boundary factor decreased. On the other hand, if the lengthscale is bigger than that generated by the figure means that the approximation should be close enough.

Figure \ref{fig3_lscale_vs_J_vs_c_part1} is useful for the user to know the suitable values for the number of basis functions $m$ and the boundary factor $c$ to approximate certain process characterized by its lengthscale. This Figure will allow us to optimize computations which depends on the number of basis functions. Starting from some guess of the lengthscale of the process and knowing the range of the input data which we are interested in, the user can do a few iterative adjustments to obtain an optimal fit. This figure also provides a diagnosis tool of the fit by comparing a actual estimated lengthscale to that the figure bring.

@@

If we look back to the conclusions drawn from Figures \ref{fig5_MSE_vs_J} and \ref{fig6_lscale_vs_J} where 10 basis functions and a boundary factor of 1.5 would be enough to closely approximate a function with normalized lengthscale equal to 0.3, we can recognize that these conclusions also matches the conclusion brought from Figure \ref{fig3_lscale_vs_J_vs_c_part1}.

\subsection{Comparative analysis between the lengthscales estimated by a regular GP model and approximate GP model}

We make a comparative analysis between the lengthscales estimated by the regular GP model and the approximate GP model in several datasets with different "wigglyness" (or lengthscales). Basically, comparing the lengthscales we can recognize if an approximation is close enough to the regular GP. We contrast these results with the information brought from Figure \ref{fig3_lscale_vs_J_vs_c_part1}.

For this analysis, we fit the regular GP model and the approximate GP model on different datasets. These datasets consists of noisy draws from a GP prior model with a squared exponential covariance function with different lengthscale. Different number of basis functions are used to fit the approximate GP model. The boundary factor is set in every case to a right and optimal choice. Figure \ref{fig7_varing_lscale} shows the posterior means of the regular GP model and the approximate GP model with different number of basis functions, for every dataset.

@@

The comparative analysis between the lengthscales, estimated by a regular GP model and an approximate GP model in the different datasets with different lengthscales, is shown in Figure \ref{fig8_Tlscale_vs_Elscale}. In the X-axes of the plots, the true lengthscales of the data-generative functions for every dataset scenario are represented. In the Y-axes, the estimated lengthscales for the regular GP and approximate GP models are represented. Different comparisons are made using different number of basis functions in the approximate GP model. Let's assume that the boundary factor was properly set in every case. The dashed black line represents the minimum lengthscales that can be closely approximate given the number of basis functions and the boundary factor, according to Figure \ref{fig3_lscale_vs_J_vs_c_part1}. Figure \ref{fig9_MSE_varing_lscale} shows the SRMSE of the approximate GP model, computed against the regular GP model, as a function of the lengthscale and number of basis functions.

@@

>>>>

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the non-linearity of the estimated function, which is in turn characterized by the lengthscale of the covariance function.
In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

% This graph basically provides the recommended values for these factors in order to get a close approximation. This graph also allows us to diagnose an actual fit by checking whether the value of the estimated lengthscale does not keep to those that the Figure brings as suitable ones.

%We compare regular GP and HSGP fitted for different dataset with different "wigglyness".
%The comparison is made through comparing their estimated lengthscales.

@@

As explained in Section 2, the approximation of the covariance function is a series expansion of eigenfunctions and eigenvalues of the Laplace operator in a given domain $\Omega$, for instance in a one-dimensional input space $\Omega=[-L,L]\subset \mathbb{R}$:
%
\begin{equation}\label{diffcov}
k(\tau) = \sum_{j=1}^{\infty} S_{\theta} \left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0),  \nonumber
\end{equation}

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equation (\ref{bf&lambda}) for the unidimensional case and in Equation (\ref{bf&lambda_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})), which will affect its accuracy especially near the boundaries. As we wil see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

\noindent where $S$ (for $S > 0$) represents the half-range of the input space, and $c$ (for $c \geq 1$) is the proportional extension factor. In the following, we will to refer to $c$ as the boundary factor of the approximation. The boundary factor can also be regarded as the boundary $L$ normalized by the half-range $S$ of the input space.

We start with an illustration on how the number of basis functions $m$ and boundary factor $c$ influences the accuracy of the HSGP approximations, separately. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$ using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the non-linearity of the estimated function: Fewer basis functions imply smoother covariance functions (Paul: when would you consider a covariance function as smooth? Also, I think the main point here is that at long distances the implied covariance function becomes negative which can't be in an exact GP model) and consequently more linear functional forms. The higher the "wigglyness" of the function to be estimated, the more basis functions will be required (Paul: do we see this in Figure \ref{fig1_Post_J} already?). For $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries as well as covariances at long distances.

@@

Next, we will focus on analyzing the interaction effects between these $m$ and $c$ on the performance of the approximation. The lengthscale and marginal variance will no longer be fixed but rather estimated in both regular GP and HSGP models.  
%The right-side plots in Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} are computed using the lengthscale and magnitud values of the true GP prior. Now, we focus on the estimated lengthscale and magnitud parameters after fitting the model, either of the regular GP model fit and approximate GP model fit. 
Figure \ref{fig4_Post_&_Cov_part1} shows the functional posterior predictions and the covariance function obtained after fitting the data, for varying $m$ and $c$. Figure \ref{fig5_MSE_vs_J} shows the root mean square error (RMSE) of the approximate GP model, computed against the regular GP model. Figure \ref{fig6_lscale_vs_J} shows the estimated lengthscale and marginal variance for the regular GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig5_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computations would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the choice of $m = 10$ and $c = 1.5$ could still be an accurate enough choice. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure \ref{fig4_Post_&_Cov_part1}. From these results, some general conclusions may be drawn:

\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa).
\item There exists a minimum $c$ under which a close approximation will never be achieved regardless of $m$.
\end{itemize}

@@

Additionally, there is a clear relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figure \ref{fig3_lscale_vs_J_vs_c_part1} depicts how these three factors interact with each other in relation to the performance of the HSGP approximation (Paul: Did you say how you choose $\varepsilon$?). On the X-axis of this plot is placed the lengthscale of the process normalized by the half-range of the data $\ell/S$. The countour lines gather the boundary factor $c$. And the Y-axis represents the number of basis functions (Paul: This needs to be put in the figure caption not in the text). More precisely, for a given GP function with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig3_lscale_vs_J_vs_c_part1} shows the minimum $m$ required to achieve a close approximation in terms of satisfying Equation (\ref{diff_covs}). Alternatively, this figure could be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be closely approximated given $m$ and $c$. We obtain the following main conclusions:

\begin{itemize}
\item As $\ell$ increases, $c$ and $m$ required for a close enough approximation decrease.
\item There is a minimum $c$ under which a close approximation is never going to be achieved (The contour lines have an end in function of the lengthscale; Paul: what do you mean?).
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve a close approximation.
\end{itemize}

For users of HSGP models, Figure \ref{fig3_lscale_vs_J_vs_c_part1} also severs as a powerful diagnostic tool in determining if the obtained accuracy is acceptable.  Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ is exceeds the corresponding line provided in Figure \ref{fig3_lscale_vs_J_vs_c_part1}. If $\hat{\ell}$ exceeds the line, the approximation should be close enough. If, however, it does not exceed it, the approximation may be inaccurate and $m$ should be increased or $c$ decreased (Paul: decreasing $c$ seems dangerous to me if predictions for out of sample data are desired. Shell we mention this?). We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can fit an HSGP model, check the approximation accuracy, and, if not accurate enough, repeat the process while replacing $\ell$ by the newly obtained estimate $\hat{\ell}$.

@@

If we look back to the conclusions drawn from Figures \ref{fig5_MSE_vs_J} and \ref{fig6_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were enough to closely approximate a function with $\ell = 0.3$, we can recognize that these conclusions also matches those obtained from Figure \ref{fig3_lscale_vs_J_vs_c_part1}.

\subsection{Comparing lengthscale estimates}

Another way to look at the approximation accuracy of HSGPs is to compare the lengthscale estimates
obtained from the regular GP and HSGP models. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well.

For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function (Paul: Did we talk about the applied kernels previously in Section 3?) and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models while the boundary factor $c$ is fixed to some reasonable value. Figure \ref{fig7_varing_lscale} shows the posterior predictions of the regular GP and HSGP models fitted to those datasets.

@@

The Lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. In the X-axes of the plots, the true lengthscales of the data-generative functions for every dataset scenario are represented. In the Y-axes, the estimated lengthscales for the regular GP and approximate GP models are represented (Paul: This needs to go into the Figure captions). The dashed black line represents the minimum lengthscales that can be closely approximate given the number of basis functions and the boundary factor, according to Figure \ref{fig3_lscale_vs_J_vs_c_part1} (This also needs to go into the Figure caption, but may be worth keeping in the text as well). Further, Figure \ref{fig9_MSE_varing_lscale} shows the SRMSE (Paul: you did not define that anywhere before) of the HSGP against the regular GP model as a function of the lengthscale and number of basis functions.

@@

(Paul: I don't get the last paragraph. Can you rewrite this a little bit yourself and also merge it with the former paragraph? I am happy to look at it again afterwards.)
